---
title: "Running and fitting models on naturalistic corpora"
author: "George Kachergis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Running and fitting models on naturalistic corpora}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(XSLmodels)

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

Most of `XSLmodels` is built around *controlled* cross-situational learning
experiments: a handful of words and objects, a designed training order, and a
vector of human accuracies to fit with sum-of-squared-error. This vignette is
about the other kind of input -- **naturalistic corpora** of caregiver speech
paired with the objects present in the scene. There is no human
referent-selection data to fit; instead we ask how well a model's learned
word--object matrix recovers a **gold-standard lexicon**.

Two such corpora ship with the package, imported from the
[wurwur](https://github.com/mcfrank/wurwur) package:

```{r}
rollins_corpus$data
fm_corpus$data
```

- **`rollins_corpus`** -- 619 mother-to-infant utterances from the Rollins
  corpus in CHILDES; the corpus the intentional Bayesian model of Frank,
  Goodman & Tenenbaum (2009) was originally fit to.
- **`fm_corpus`** -- the larger Frank, Tenenbaum & Fernald corpus: 4763
  utterances, and (unlike Rollins) the speaker's referential intention is
  hand-coded for each utterance.

Each is a list, not a bare `xslData`:

```{r}
names(rollins_corpus)
names(fm_corpus)
```

`$data` is the training `xslData`; `$gold` is the reference lexicon. Note that
words and objects here are the actual strings, not integer ids:

```{r}
rollins_corpus$data$train$words[[1]]
rollins_corpus$data$train$objects[[1]]
str(rollins_corpus$gold)
```

## Running a model

`xsl_run()` works exactly as it does for experimental conditions -- it returns
a fitted object whose `matrix` is the learned word (rows) by object (columns)
association matrix.

```{r}
res <- suppressWarnings(xsl_run(baseline(), rollins_corpus$data))
m <- res$fits[[1]]$matrix
dim(m)

# co-occurrence counts for one word (this corpus is a picture-book session,
# so the objects are animals and toys -- "moocow", "piggie", etc.)
sort(m["moocow", ], decreasing = TRUE)[1:4]
```

Two things to know:

- **Ignore `res$sse` and `res$fits[[1]]$perf`.** Those compare against
  `data$accuracy`, which is empty for a corpus. `xsl_run()` still computes them
  (falling back to `get_perf()`), which is where the
  `"longer object length is not a multiple of shorter"` warning above comes
  from -- the matrix is not square. It is harmless here; we score against the
  gold lexicon instead.
- The batch model `fgt2009()` (below) does joint inference over the whole
  corpus and has no learning trajectory, so `control$reps` does nothing for it.

## Scoring against the gold lexicon

The evaluation helpers take a `gold_lexicon = list(words, objects)`. Because a
model's matrix is real-valued, the score depends on a **threshold** for
counting an edge as "learned". [get_fscore()] scores one threshold:

```{r}
get_fscore(m / rowSums(m), threshold = 0.1, gold_lexicon = rollins_corpus$gold)
```

[get_roc()] sweeps thresholds (row-normalizing the matrix first), and
[get_roc_max()] returns the best F-score over that sweep -- the standard
single-number summary:

```{r}
roc <- get_roc(m, gold_lexicon = rollins_corpus$gold)
head(roc, 3)

get_roc_max(m, gold_lexicon = rollins_corpus$gold)
plot_roc(m, gold_lexicon = rollins_corpus$gold)
```

Comparing a few associative models on Rollins:

```{r}
assoc_models <- list(
  baseline        = baseline(),
  decay           = decay(C = 0.99),
  rescorla_wagner = rescorla_wagner(C = 1, alpha = 0.1, beta = 0.1, lambda = 1),
  fazly           = fazly(lambda = 1e-5, beta = 8500)
)

sapply(assoc_models, function(mod) {
  mm <- suppressWarnings(xsl_run(mod, rollins_corpus$data)$fits[[1]]$matrix)
  round(get_roc_max(mm, gold_lexicon = rollins_corpus$gold), 3)
})
```

## The intentional Bayesian model

`fgt2009()` is the model built for exactly this problem: it does joint
posterior inference over a word--object lexicon for the whole corpus at once,
marginalizing the speaker's referential intention on each utterance. It is a
**batch** MCMC model, so a single run takes seconds to a minute or two on these
corpora (much slower than the associative tallies) and `xslFit$matrix` holds
posterior edge marginals `P(word names object | corpus)`.

**The `gamma` parameter matters here.** `gamma` is the probability a word is
used *referentially*. In a controlled experiment every word is a label, so the
default is `gamma = 1`. Naturalistic speech is full of non-referential words
("the", "look", "you", "see"), so a corpus wants the published
`gamma = 0.1`, `kappa = 0.05` -- otherwise the model is forced to explain
every function word as naming something and learns almost nothing.

```{r fgt-run, eval = FALSE}
res <- xsl_run(
  fgt2009(alpha = 3, gamma = 0.1, kappa = 0.05),
  rollins_corpus$data
)
m_fgt <- res$fits[[1]]$matrix

# the learned lexicon: edges with posterior probability > 0.5
edges <- which(m_fgt > 0.5, arr.ind = TRUE)
data.frame(word   = rownames(m_fgt)[edges[, 1]],
           object = colnames(m_fgt)[edges[, 2]],
           prob   = round(m_fgt[edges], 2))
#>       word object prob
#> 1     bear   bear 1.00     # correct
#> 2   bottle   bear 1.00     # "baby bottle" co-occurs with the bear
#> 3  bigbird   bird 1.00     # correct
#> 4     book   book 1.00     # correct
#> 5   moocow    cow 1.00     # correct
#> 6      hat    hat 1.00     # correct
#> 7     oink    pig 1.00     # correct
#> ...                        # 28 edges total

get_roc_max(m_fgt, gold_lexicon = rollins_corpus$gold)
#> [1] 0.454   # MCMC: expect a little run-to-run variation
```

## "Fitting": sweeping the lexicon-size prior

`fgt2009()`'s `alpha` penalizes lexicon size (`P(L)` proportional to
`exp(-alpha * |L|)`) and has to scale with corpus size: too small gives a
diffuse over-large lexicon, too large an empty one. The F-vs-`alpha` curve is
single-peaked, so the right way to "fit" the model to a corpus is a coarse
sweep -- which is what [fgt2009_sweep_alpha()] does. Pass `gold` and it reports
the best-threshold F-score at each `alpha`:

```{r sweep, eval = FALSE}
sweep <- fgt2009_sweep_alpha(
  rollins_corpus$data,
  alphas = c(2, 4, 6, 9, 14),
  gold   = rollins_corpus$gold,
  gamma = 0.1, kappa = 0.05,
  n_warmup = 50, n_samples = 120      # a lighter sampler for the sweep
)
sweep
```

```{r sweep-precomputed, echo = FALSE}
# precomputed -- each row is a full MCMC run (~40s); numbers wobble a little
# from run to run because the sampler is stochastic.
sweep <- data.frame(
  alpha = c(2, 4, 6, 9, 14),
  f_max = c(0.429, 0.404, 0.390, 0.400, 0.290)
)
plot(sweep$alpha, sweep$f_max, type = "b", pch = 19,
     xlab = expression(alpha), ylab = "best-threshold F",
     main = "fgt2009 on the Rollins corpus (gamma = 0.1)")
```

The curve rises to a shallow peak around `alpha = 2`--`4` and then falls off
as the prior forces the lexicon empty. Take the `alpha` near the peak; that is
the value to report, and the one to use for trial-by-trial predictions
([predict_referent()]) or further analysis. On this small corpus `fgt2009` is
in the same range as the better associative models (`fazly` above); its
advantage is largest on data with more varied, informative co-occurrence.

## The FM corpus and its coded intentions

`fm_corpus` is used the same way, but it is bigger and comes with extra
structure. `fm_corpus$intents[[t]]` is the object(s) the speaker actually
referred to on utterance `t` -- empty for the roughly half of utterances that
are non-referential:

```{r}
fm_corpus$data$train$words[[1]]
fm_corpus$intents[[1]]

mean(lengths(fm_corpus$intents) == 0)   # fraction non-referential
```

It also ships three gold lexicons -- a hand-curated one plus two auto-derived
from the coded intentions at different co-occurrence thresholds:

```{r}
lengths(lapply(list(curated    = fm_corpus$gold,
                    strict     = fm_corpus$gold_variants$strict,
                    permissive = fm_corpus$gold_variants$permissive),
               `[[`, "words"))
```

```{r fm-score}
m_fm <- suppressWarnings(xsl_run(baseline(), fm_corpus$data)$fits[[1]]$matrix)

sapply(list(curated    = fm_corpus$gold,
            strict     = fm_corpus$gold_variants$strict,
            permissive = fm_corpus$gold_variants$permissive),
       function(g) round(get_roc_max(m_fm, gold_lexicon = g), 3))
```

The permissive lexicon is larger and easier to hit some of, so absolute
F-scores are not comparable across the three -- pick one and use it
consistently when comparing models.

## A quick model comparison

`tests/bakeoff_comparison/corpus_gold_comparison.R` runs every model once on
each corpus and scores the learned matrix against the corpus gold lexicon.
It is deliberately **not** a fitted bake-off: the associative and sampling
models use `xsl_model_registry()`'s starting parameters (calibrated defaults,
not re-fit here), and `fgt2009()` uses `gamma = 0.1`, `kappa = 0.05` with
`alpha` from the two-point heuristic `round(7 * sqrt(n / 600))`.

```{r comparison-table, echo = FALSE}
# precomputed by tests/bakeoff_comparison/corpus_gold_comparison.R
cmp <- data.frame(
  model = c("bayesian_decay", "fazly", "fgt2009_rsa", "fgt2009",
            "uncfam_sampling", "multi_sampling", "uncfam_attention", "uncfam",
            "guess_and_test", "softmax_rl", "baseline", "rescorla_wagner",
            "decay", "propose_but_verify", "kalman_filter", "pursuit",
            "uncfam_predictive", "tilles"),
  Rollins = c(0.449, 0.450, 0.395, 0.385, 0.377, 0.377, 0.371, 0.371,
              0.409, 0.407, 0.377, 0.312, 0.320, 0.321, 0.303, 0.353,
              0.208, 0.080),
  FM = c(0.450, 0.407, 0.421, 0.421, 0.353, 0.353, 0.354, 0.353,
         0.297, 0.258, 0.268, 0.290, 0.272, 0.236, 0.211, 0.048,
         0.091, 0.048)
)
knitr::kable(cmp, digits = 3,
             caption = "Best-threshold F of the learned lexicon vs. the gold lexicon.")
```

Reading the table:

- **`bayesian_decay` and `fazly` come out on top** (F around 0.45 on both
  corpora), with `fgt2009` close behind.
- **`fgt2009` trades recall for precision.** On FM its precision is ~0.71
  against ~0.2--0.34 for the others: the geometric size prior makes it commit
  to a small, mostly-correct lexicon rather than a large noisy one. Whether
  that is the right tradeoff depends on what you want the lexicon for.
- The RSA layer makes no difference on FM (`fgt2009` == `fgt2009_rsa`): at
  `alpha = 20` the learned lexicon is sparse enough that no word names two
  co-present objects, so the pragmatic speaker reduces to the literal one.
- The middle of the pack is a tight cluster around F = 0.35 -- the three
  `uncfam` variants and the two sampling versions all land there, which is
  reassuring since the sampling models are meant to approximate `uncfam()`.
- **Every model now runs on both corpora**, but several needed fixes to get
  there. The sampling/RL models were built for controlled experiments of a
  few dozen trials; running them on 4763 naturalistic utterances exposed a
  memory blow-up (`xslControl()`'s `keep_traj = FALSE` default), an error on
  the many non-referential utterances with no objects present, and two
  latent bugs (`softmax_rl()` and `guess_and_test()` comparing a sampled
  guess to object *positions* vs *labels*; `guess_and_test()` also mishandling
  a word repeated within one utterance; `tilles()` coercing its labels with
  `as.integer()` and returning an unnamed matrix). `pursuit` and `tilles`
  still learn near-empty or diffuse lexicons on FM at the registry's default
  parameters -- read the low scores as "this parameterisation doesn't
  transfer", not "the model is broken".

### Fitting to a corpus

`tests/bakeoff_comparison/corpus_gold_fits.R` goes a step further: for each
model it runs a (parallel) DEoptim search over the model's parameters,
maximizing the same best-threshold F against the gold lexicon (these corpora
have no accuracy vector, so `xsl_fit()`'s SSE objective does not apply).
`fgt2009()` is fit by an `alpha` sweep instead. `uncfam_sampling()` /
`multi_sampling()` are left at defaults -- averaging many simulations per
DEoptim evaluation over 4763 utterances is impractical.

```{r fits-table, echo = FALSE}
# precomputed by tests/bakeoff_comparison/corpus_gold_fits.R
fit <- data.frame(
  model = c("bayesian_decay", "fgt2009", "fazly", "softmax_rl",
            "uncfam_attention", "uncfam", "guess_and_test", "tilles",
            "uncfam_predictive", "propose_but_verify", "rescorla_wagner",
            "decay", "kalman_filter"),
  Rollins_default = c(0.449, NA, 0.450, 0.361, 0.371, 0.371, 0.378, 0.080,
                      0.208, 0.310, 0.312, 0.320, 0.303),
  Rollins_fitted  = c(0.465, 0.452, 0.468, 0.449, 0.438, 0.422, 0.440, 0.417,
                      0.360, 0.365, 0.348, 0.382, 0.382),
  FM_default = c(0.450, NA, 0.407, 0.276, 0.354, 0.353, 0.308, 0.048,
                 0.091, 0.246, 0.290, 0.272, 0.211),
  FM_fitted  = c(0.560, 0.529, 0.412, 0.463, 0.443, 0.446, 0.362, 0.385,
                 0.411, 0.326, 0.324, 0.288, 0.274)
)
knitr::kable(fit, digits = 3,
             caption = "Best-threshold F: registry default vs. fit to the corpus.")
```

- **Fitting lifts every model**, and by a lot for the ones whose registry
  defaults were tuned on controlled experiments: `tilles` (+0.34 on both
  corpora), `uncfam_predictive` (+0.15 / +0.32), `softmax_rl` (+0.09 / +0.19).
- **`bayesian_decay` fit to FM reaches F = 0.56** -- clearly the strongest
  result -- with `fgt2009` next at 0.53. `fazly` barely moves (its defaults
  were already near-optimal for these corpora).
- The fitted parameters are in `corpus_gold_fits.rds`; a few sit near a
  registry bound (`kalman_filter`'s `sigma2_obs` fits to ~498 against an upper
  bound of 500), a hint that those bounds could be widened for corpus-scale
  data.

## References

Frank, M. C., Goodman, N. D., & Tenenbaum, J. B. (2009). Using speakers'
referential intentions to model early cross-situational word learning.
*Psychological Science, 20*(5), 578--585.

Frank, M. C., Tenenbaum, J. B., & Fernald, A. (2013). Social and discourse
contributions to the determination of reference in cross-situational word
learning. *Language Learning and Development, 9*(1), 1--24.
