---
title: "Importing a Dataset and Fitting Models"
author: "George Kachergis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Importing a Dataset and Fitting Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
library(XSLmodels)

knitr::opts_chunk$set(
  eval = FALSE, # temporary until vignette updated
  collapse = TRUE,
  comment = "#>"
)
```

Whether you already have human data from an experiment that you would like to fit models to, or if you would like to simply evaluate model performance on a given experiment's training trials, this vignette will show to 1) structure training trials, 2) examine the structure of an experiment, 3) run model(s) through the training trials, and 4) specify how the model(s) should be tested.

## Importing a Dataset

Cross-situational word learning (XSL) datasets are defined with the `xslData` class, which require defining minimally a training phase and a testing phase.
Each trial of the training phase presents participants (or models) with one or more words and one or more referents (i.e. objects).
In this package, each word $w$ and object $o$ is referred to by index (i.e., $w_i$, $o_j$), and the to-be-learned word-object pairs are assumed to have the same index (i.e. $w_i = o_i$).
Thus, each dataset defines a list `train` of `words` and `objects` indices on each trial, also defined in lists (to allow for a potentially varying number of words/objects per trial).
The `test` trials are defined by a list of the available objects on each test trial
A dataset can also be given a short `label`, and a longer (maybe per-item?) `condition`.

If you already have collected a sample of human participants, you can optionally include per-item `accuracy`, a vector of length = number of to-be-learned items) and the number of participants (`n_subj`).
This sample size and accuracy can then be compared to model performance, and if desired, used to fit model parameters.

Below we show how to define a new dataset.

```{r, echo=T}
xslData(train = list(words = list(c(1, 2), 
                                  c(2, 3)),
                     objects = list(c(1, 2),
                                    c(2, 3))),
        test = list(1:3, 1:3, 1:3),
        accuracy = c(.3, .7, .5)) # length(unique(unlist(x$train$words))) == length(x$accuracy) is not TRUE
# need to define accuracy -- should we make it optional?
# why does this result in "test trials: 0" ?
```


## 1. Structure of an experimental condition

get_example_ambiguous_condition() and get_example_unambiguous_condition() return simple example experimental conditions, which consist of a list of training *trials*, an optional vector *perf* containing the per-word performance of people on the test trials, and an optional list of test trials, enumerating each to-be-tested word and the set of referents to be presented for choosing among. 
The list of training trials is further composed of a list of the words presented on each trial, and a list of the referents (*objs*) presented on each trial.

```{r, echo=T}
ag <- XSLmodels::get_example_ambiguous_condition()

ag
```

Shown above, the example ambiguous condition has three training trials (`ag$trials`). 
On the first trial words 1 and 2 appear (`ag$trials$words[[1]]`) alongside referents 1 and 2 (`ag$trials$objs[[1]]`; order of words/referents within a trial is irrelevant).

## 2. Viewing an experiment's structure

One common way to summarize the structure of a cross-situational word learning experiment is to make a matrix of the word-object co-occurrences, which can be visualized with a heatmap to show the strength of association between each word and object.
create_cooc_matrix() will return a word x object matrix showing the tallied co-occurrences of each word with each object across all of the training trials.

```{r}
XSLmodels::create_cooc_matrix(ag$trials)
```


## 3. Run model(s) through a set of training trials

```{r run-models}
xsl_run(kachergis(X = .1, C = 1, B = .98), data = ag)
```


## 4. Evaluating model performance

Models return both a word-object matrix of associations (or 'hypotheses', i.e. binary-valued associations), as well as the conditional probability of selecting the intended referent, given each word (i.e., P(referent | word)).

```{r}
## TESTING
#pv = run_stochastic_model(conds, "trueswell2012", c(.1, .5)) # SSE=1.18
#gt = run_stochastic_model(conds, "guess-and-test", c(.1, .5)) # SSE=1.14
#pt = run_stochastic_model(conds, "pursuit_detailed", c(.2, .3, .05)) # SSE=4.06

#tl = fit_model("tilles", conds[[1]], c(1e-5,1e-5,1e-5), c(1,1,1)) # run_model



# temporary (until we finish full fits)
#group_fits[["trueswell2012"]] = list(pars = c(0.113666, 0.266792)) # SSE=.879
#group_fits[["guess-and-test"]] = list(pars = c(0.691312, 0.991726)) # SSE=.884
#group_fits[["pursuit_detailed"]] = list(pars = c(0.088324, 0.409523, 0.000059)) # SSE = 0.996
#group_fits[["kachergis_sampling"]] = list(pars = c(0.15553, 11.04866, 0.99260)) # SSE = 0.407
#gfd = get_model_dataframe(group_fits, combined_data) # 726 items per model
#save(group_fits, gfd, file="fits/group_fits.Rdata")

```


## 5. Included experimental conditions

```{r}
# A dataset of `r length(names(XSLmodels::combined_data))` experimental conditions is included in the package in XSLmodels::combined_data.
# Below, we show this data can be used to optimize parameters for existing models in the package.

assoc_model_names = c("kachergis", "fazly", "strength", "uncertainty", 
           "novelty", "Bayesian_decay", "rescorla_wagner")
stochastic_model_names = c("guess-and-test","pursuit","trueswell2012","kachergis_sampling")

```

