---
title: "Overview of `XSLmodels`: Importing cross-situational datasets and fitting models"
author: "George Kachergis & Mika Braginsky"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Importing Cross-situational Datasets and Fitting Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
library(XSLmodels)

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

The main goal of the `XSLmodels` package is to enable researchers to run and optimize a collection of cross-situational word learning (XSL) models on one or more datasets from cross-situational word learning experiments--and relevant corpora. This vignette provides an overview of the structure and definition of XSL datasets -- covering both datasets provided in the packages, and how to import your own data -- and then turns to discussing how to run, optimize, and even create your own XSL model.

# Defining and structure of cross-situational learning (XSL) datasets.

Whether you already have human data from an experiment that you would like to fit models to, or if you would like to simply evaluate model performance on a given experiment's training trials, this vignette will show to 1) structure training trials, 2) examine the structure of an experiment, 3) run model(s) through the training trials, and 4) specify how the model(s) should be tested.

## 0. Importing a Dataset

Cross-situational word learning (XSL) datasets are defined with the `xslData` class, which requires defining minimally a training phase and a testing phase.
Each trial of the training phase presents participants (or models) with one or more words and one or more referents (i.e. objects).
In this package, each word $w$ and object $o$ is referred to by index (i.e., $w_i$, $o_j$), and the to-be-learned word-object pairs are assumed to have the same index (i.e. $w_i = o_i$).
Thus, each dataset defines a list `train` of `words` and `objects` indices on each trial, also defined in lists (to allow for a potentially varying number of words/objects per trial).
The `test` trials are defined by a list of the available objects on each test trial.
A dataset can also be given a short `label`, and a longer (maybe per-item?) `condition`.

If you already have collected a sample of human participants, you can optionally include per-item `accuracy`, a vector of length equal to the number of to-be-learned items) and the number of participants (`n_subj`).
This sample size and accuracy can then be compared to model performance, and if desired, used to fit model parameters.

Below we show how to define a new dataset.

```{r}
xslData(train = list(words = list(c(1, 2), c(2, 3)),
                     objects = list(c(1, 2), c(2, 3))),
        test = list(words = list(1, 2, 3),
                    objects = list(1:3, 1:3, 1:3)),
        label = "example dataset",
        condition = "example condition")
```


## 1. Structure of an experimental condition

`get_example_ambiguous_condition()` and `get_example_unambiguous_condition()` return simple example experimental conditions, which consist of a list of training *trials*, an optional vector *perf* containing the per-word performance of people on the test trials, and an optional list of test trials, enumerating each to-be-tested word and the set of referents to be presented for choosing among. 
The list of training trials is further composed of a list of the words presented on each trial, and a list of the referents (*objs*) presented on each trial.

```{r}
ag <- get_example_ambiguous_condition()
ag
```

Shown above, the example ambiguous condition has three training trials (`ag$trials`). 
On the first trial words 1 and 2 appear (`ag$trials$words[[1]]`) alongside referents 1 and 2 (`ag$trials$objs[[1]]`; order of words/referents within a trial is irrelevant).

## 2. Viewing an experiment's structure

One common way to summarize the structure of a cross-situational word learning experiment is to make a matrix of the word-object co-occurrences, which can be visualized with a heatmap to show the strength of association between each word and object.
`create_cooc_matrix()` will return a word x object matrix showing the tallied co-occurrences of each word with each object across all of the training trials.

```{r}
create_cooc_matrix(ag$trials)
```

## 3. Included experimental conditions

A dataset combining `r length(xsl_datasets)` experimental conditions is included in the package in `xsl_datasets`.
For example, here is a summary of one experimental condition in the dataset: an asymmetric (3x4; i.e. 3 words and 4 objects per trial) condition with 36 training trials, 18 words and 18 objects with corresponding mean accuracies from 25 subjects.
(Note that `test trials: 0` indicates that this condition used 18AFC test trials, on which all referents are given for each tested word. Conditions which show a subset of the trained word-referent pairs (e.g. 4AFC) specify a list of referents shown on for each tested word.)

```{r, example-xsl-dataset}
xsl_datasets[[1]]
```

The properties condition can be accessed and examined in detail:

```{r}
# properties of an experimental condition:
names(xsl_datasets[[1]])
# e.g., mean accuracy for each tested word (i.e., P(choosing correct referent | word)):
xsl_datasets[[1]]$accuracy
```


# Running, fitting, and defining XSL models.

## 4. Run model with given parameters through a single dataset and pull SSE.

Now that the definition and structure of the datasets is clear, we will turn to running and fitting models, before finally discussing the structure of an `xslMod`, and how to extend an existing model or create a new one.

```{r run-models}
run1 <- xsl_run(uncfam(X = .1, C = 1, B = .98), data = xsl_datasets[[1]])
run1$sse
```

## 5. Run a model with given parameters through multiple datasets and pull SSE.

```{r}
run3 <- xsl_run(uncfam(X = .1, C = 1, B = .98), data = xsl_datasets[1:3])
run3$sse
```


## 6. Evaluating model performance

Models return both a word-object matrix of associations (or 'hypotheses', i.e. binary-valued associations), as well as the conditional probability of selecting the intended referent, given each word (i.e., P(referent | word)).

```{r}
# Test different models on a subset of datasets
gt <- xsl_run(guess_and_test(f = .1, sa = .5), data = xsl_datasets[1:3])
gt$sse

pt <- xsl_run(pursuit(gamma = .2, threshold = .3, lambda = .05), data = xsl_datasets[1:3])
pt$sse

# Compare with baseline
bl <- xsl_run(baseline(), data = xsl_datasets[1:3])
bl$sse
```

## 7. Using the helper functions

The package provides several helper functions to make it easier to work with models and datasets:

```{r}
# See what models are available
models <- show_models()
models

# See what datasets are available
datasets <- show_datasets()
head(datasets)
```

## 8. Model fitting

Below, we show this data can be used to optimize parameters for existing models in the package.

```{r}
# Fit a simple decay model to a single dataset
fit_result <- xsl_fit(decay(C = .98), data = xsl_datasets[[1]], 
                      lower = .8, upper = 1.0)
fit_result[[1]]$optim$bestmem
fit_result[[1]]$optim$bestval
```

## 9. Available models

The package includes several types of models:

**Association-based models:**
- `uncfam()` - Uncertainty and familiarity biased model
- `fazly()` - Fazly et al. model
- `rescorla_wagner()` - Rescorla-Wagner model
- `bayesian_decay()` - Bayesian decay model

**Sampling-based models:**
- `uncfam_sampling()` - Sampling version of uncfam
- `multi_sampling()` - Multi-hypothesis sampling
- `guess_and_test()` - Guess and test model
- `propose_but_verify()` - Propose but verify model
- `pursuit()` - Pursuit model

**Baseline models:**
- `baseline()` - Simple co-occurrence baseline
- `decay()` - Decay model
- `tilles()` - Tilles model

Each model can be run with different parameters and compared to human data to understand which learning mechanisms best explain cross-situational word learning behavior.
